---
title: "RNAseq pipeline - create your own gene matrix"
output: html_notebook
author: Cristina Venturini
---

This is a script and instruction to download RNAseq tpm files from the RDS server and create your own matrix. 
You need: 

- to install https://mobaxterm.mobatek.net/ if you are usinf Windows

- to install Rstudio

- a list of tpm files you are interested in: from the template, filter what you need and copy the file names into a .txt file (see example_files.txt)



To start we call an external script with some useful functions we will use later
```{r}
source("scripts/rnaseq_pipeline.r")
```


##Get files from RSD
1) Open Mobaxterm 

2) Move into the directory you want:
Instead of /home/cristina/Downloads/ you need to put the full path to the directory
cd /home/cristina/Downloads/tutorial_rnaseqpipeline-master

1) To add .txt extension to each file name run this command in mobaxterm: 
sed -i -e 's/$/.txt/g' example_files.txt

2) To download files from RSD into your own computer, run this command in mobaxterm: 
NB Replace <username> with your UCL username


rsync -av --files-from=example_files.txt <username>@ssh.rd.ucl.ac.uk:/rd/live/ritd-ag-project-rd002u-mnour10/RNAseq/TPM_files/ files_fromrsd/

You will be asked to type your UCL password

Files will be downloaded into a new folder called files_fromrsd


##Create your own matrix 

1) Each tpm file might have a different total number of genes due to different Ensembl reference used in the mapping process. We will extract genes that are present in all files. 
You need to create a new folder for the files, let's call it "files_fromrsd_equal" (if you are feeling super confident with Mobaxterm, you can do in the terminal: mkdir files_fromrsd_equal )

Run this to equalise the number 
```{r}
input= "files_fromrsd/" ##CHANGE HERE
output = "files_fromrsd_equal/"  ##CHANGE HERE
print("Did you remember to create the output directory?")
Equaliser(input =input, output= output)
```




2) Let's put the files together in a matrix 
```{r}

#install.packages(c("dplyr", "tidyr","data.table"), dependencies = T)
library(plyr)
library(dplyr)
library(tidyr)
#require("tidyverse")
library(data.table)

paths <- dir("files_fromrsd_equal/", pattern = "\\.txt$", full.names = TRUE)
names(paths) <- gsub("\\.txt", "", dir("files_fromrsd"))
tpm_df <- ldply(paths, fread, stringsAsFactors = FALSE,header=FALSE) %>% spread(key=.id,value=V2) 
row.names(tpm_df) <- tpm_df$V1
tpm_df$V1 <- NULL
head(tpm_df)
```


3) Expression values < or = to 0.001 are transformed into 0.001 to avoid problems with log2 transformation. We then log2 normalise all data.

We write a .csv file with this data, change the file name as you like (here is example_tpm_log2_0.001.csv)


```{r}
tpm_df[tpm_df <= 0.001] <- 0.001
tpmlog <- log2(tpm_df)
tpmlog$ensembl_gene_id <- row.names(tpmlog)
tpmlog <- tpmlog %>% select(ensembl_gene_id,everything()) ##make sure biomart is not loaded, it creates problems with dplyr
write.csv(tpmlog,"example_tpm_log2_0.001.csv",quote=FALSE,row.names = FALSE) ##CHANGE HERE the output file name
```


4) We can annotate the Ensembl gene ID to gene names 
fileName is the .csv file we saved above (example_tpm_log2_0.001.csv)
If needed you can change the species for annotation: hsapiens_gene_ensembl is human 
This command will save another csv file called Annotatedexample_tpm_log2_0.001.csv
```{r}
## try http:// if https:// URLs are not supported
#source("https://bioconductor.org/biocLite.R")
#biocLite("biomaRt")
library(biomaRt) #important to load it after
fileName <- "example_tpm_log2_0.001.csv"
annotation("hsapiens_gene_ensembl",fileName)
```
5) We have duplicated gene names due to alternative sequences and error in the reference assembly. 
We can deduplicate the file, taking the gene with the highest mean and getting rid of all duplicates. 
```{r}
detach("package:biomaRt", unload=TRUE)
dedup_genesymbol("Annotatedexample_tpm_log2_0.001.csv","Annotated_example_tpm_log2_0.001_dedup.csv")
```


6) We can create a density plot of our data to help us identify outlier
Two options:
- simple plot
- interactive plot 

For the simple plot run: 
```{r}
##SIMPLE PLOT
##install.packages("ggplot2","reshape")
library(ggplot2)
library(reshape)
data_melted <- melt(tpmlog)
p <- ggplot(data_melted,aes(x=value, col= variable)) + geom_density() + theme_bw() + theme(legend.position = "None")
plot(p)

##if you want to save the plot in a pdf file - uncomment the following
#pdf("density_plot.pdf")
#plot(p)
#dev.off()
```




```{r}
##INTERACTIVE PLOT 
#install.packages("plotly")
#for for dev version
#devtools::install_github("ropensci/plotly")

library(plotly)
data_melted <- melt(tpmlog)
p <- ggplot(data_melted,aes(x=value, col= variable)) + geom_density() + theme_bw() + theme(legend.position = "None")
ggplotly(p)
```




